---
title: "NBA Field Goals"
author: "Jason Katz"
date: "12/16/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(dplyr)
library(data.table)
library(datasets)
library(ggplot2)
library(graphics)
library(grDevices)
library(lattice)
library(lubridate)
library(manipulate)
library(MASS)
library(Matrix)
library(methods)
library(mosaic)
library(mosaicData)
library(psych)
library(stats)
library(stringr)
library(utils)
library(zoo)
library(caret)
library(class)
library(rpart)
library(e1071)
library(klaR)
library(mlbench)
library(randomForest)
library(Boruta)
library(rjson)
library(grid)
library(jpeg)
library(RCurl)
library(jpeg)
library(RColorBrewer)
library(gsheet)
library(gplots)
library(viridis)
```

#Visualization of field goal percentage by zones
```{r}
#Read in the data
data <- as.data.frame(gsheet2tbl('https://docs.google.com/spreadsheets/d/1BZwnqXGPS-cJW_J1gXK22lXHRmuyaeHcdMhiqFVFQWI/edit#gid=1869485836'))

#Fix class of made basket for later analysis
data$made <- as.factor(data$made)

#Create new class of made basket for later computation
data$made_numeric <- as.numeric(data$made)-1

#Create new variable for horizontal distance to rim
data$shot_x_abs <- abs(data$shot_x)

#Create new variable for total distance from rim
data$shot_distance <- (data$shot_x^2 + data$shot_y^2)^.5

#Create new variable to categorize a shot as a two or a three
data$shot_type <- as.factor(ifelse(data$shot_y < 9.25 & data$shot_x_abs > 22, 3, ifelse(data$shot_y > 9.25 & data$shot_distance > 23.75, 3, 2)))

#Create a new variable to assign a points value to each shot (0, 2, 3)
data$points <- (as.numeric(data$made)-1)*(as.numeric(data$shot_type)+1)

#Create buckets for different range of defender distance
data$closest_defender <- as.factor(ifelse(data$defender_distance >= 0 & data$defender_distance < 1, "0-1",ifelse(data$defender_distance >= 1 & data$defender_distance < 2, "1-2", ifelse(data$defender_distance >= 2 & data$defender_distance < 3, "2-3", ifelse(data$defender_distance >= 3 & data$defender_distance < 4, "3-4", ifelse(data$defender_distance >= 4 & data$defender_distance < 5, "4-5", ifelse(data$defender_distance >= 5 & data$defender_distance < 6, "5-6", ifelse(data$defender_distance >= 6 & data$defender_distance < 7, "6-7", ifelse(data$defender_distance >= 7 & data$defender_distance < 8, "7-8", ifelse(data$defender_distance >= 8 & data$defender_distance < 9, "8-9", ifelse(data$defender_distance >= 9 & data$defender_distance < 10, "9-10",  ifelse(data$defender_distance >= 10, "10+", "NA"))))))))))))

#Create six different shot zones based on shot coordinates and other locatiom variables
data$shot_zone <- as.factor(ifelse(data$shot_type == 3 & data$shot_y < 9.25 & data$shot_x > 0, "Right Corner Three", ifelse(data$shot_type == 3 & data$shot_y < 9.25 & data$shot_x < 0, "Left Corner Three", ifelse(data$shot_type == 3 & data$shot_y > 9.25 & data$shot_y < 42.25, "Above the Break Three", ifelse(data$shot_x_abs > 8 & data$shot_type == 2 | data$shot_y > 14.25 & data$shot_type == 2, "Mid-Range", ifelse(data$shot_distance < 4, "Restricted Area", ifelse(data$shot_x_abs < 8 & data$shot_y < 14.25 & data$shot_distance > 4, "Paint (Not RA)", "Backcourt")))))))

#Upload half court image from internet
courtImg.URL <- "http://www.sawyoo.com/postpic/2011/05/nba-basketball-court-dimensions_97440.jpg"

#Save image
court <- rasterGrob(readJPEG(getURLContent(courtImg.URL)), width=unit(1,"npc"), height=unit(1,"npc"))

#Create new dataframe for only shots in the frontcourt
data_half_court <- data[which(!data$shot_zone=='Backcourt'), ]

#Create new dataframe to calculate statistics for each shot zone
data_shots <- ddply(data_half_court, .(shot_zone), summarize, shots_attempted = length(made), shots_made = sum(made_numeric))

#Calculate field field goal percentage for each shot zone
data_shots$field_goal_percentage <- (data_shots$shots_made / data_shots$shots_attempted)

#Fix format
data_shots$field_goal_percentage <- paste(as.character(round(100 * data_shots$field_goal_percentage, 1)), "%", sep="")

#Categorize each shot zone as either a two or three (for further calculations)
data_shots$type <- ifelse(data_shots$shot_zone == "Above the Break Three", 3, ifelse(data_shots$shot_zone == "Left Corner Three", 3, ifelse(data_shots$shot_zone == "Right Corner Three", 3, 2)))

#Calculate point per shot for each shot zone
data_shots$points_per_shot <- round(data_shots$type*data_shots$shots_made/data_shots$shots_attempted, 2)

#Calculate effective field goal percetange for each shot zone
data_shots$eFG <- round(ifelse(data_shots$type == 2, data_shots$shots_made/data_shots$shots_attempted, 1.5*data_shots$shots_made/data_shots$shots_attempted), 3)

#Fix format
data_shots$eFG <- paste(as.character(round(100 * data_shots$eFG, 1)), "%", sep="")

#Location for graphical purposes
data_shots$x <- ifelse(data_shots$shot_zone == "Above the Break Three", 12, ifelse(data_shots$shot_zone == "Left Corner Three", 21, ifelse(data_shots$shot_zone == "Right Corner Three", -21, ifelse(data_shots$shot_zone == "Mid-Range", -13, ifelse(data_shots$shot_zone == "Paint (Not RA)", 0, 0)))))

#Location for graphical purposes
data_shots$y <- ifelse(data_shots$shot_zone == "Above the Break Three", 25, ifelse(data_shots$shot_zone == "Left Corner Three", 0, ifelse(data_shots$shot_zone == "Right Corner Three", 0, ifelse(data_shots$shot_zone == "Mid-Range", 10, ifelse(data_shots$shot_zone == "Paint (Not RA)", 12, 3)))))

par(xpd=TRUE)

#Graph showing field goal percentages of each shot zone
ggplot(data_shots, aes(x=x, y=y)) + annotation_custom(court, -25, 25, -4.75, 42.25) + geom_point(aes(colour = shot_zone, alpha = .8), size = 50) + geom_text(aes(colour = shot_zone, label = data_shots$field_goal_percentage), vjust = .2, size = 12, color = "black", fontface = "bold") + geom_text(aes(colour = shot_zone, label = paste(data_shots$shots_made, "of", data_shots$shots_attempted, sep = " ")), vjust = 3, hjust = .5, size = 5, color = "black") + guides(alpha = FALSE, size = FALSE, colour = guide_legend(override.aes = list(size=5, alpha = 1))) + xlim(-25, 25) + ylim(-4.75, 42.25) + coord_fixed() + theme(line = element_blank(), axis.title.x = element_blank(), axis.title.y = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank(), legend.title = element_blank(), legend.text=element_text(size = 14), legend.position = c(0.2, 0.775), legend.background = element_rect(fill=0), legend.key.size = unit(1.2, "cm"), plot.title = element_text(size = 17, lineheight = 1.2, face = "bold")) 
```

#Heatmap of field goal percentage by zones and defender distance
```{r}
#Create new dataframe for further analysis
data_half_court_4_zones <- data_half_court

#Create four shot zones
data_half_court_4_zones$shot_zone <- as.factor(ifelse(data_half_court_4_zones$shot_type == 3, "Three", ifelse(data_half_court_4_zones$shot_x_abs > 8 & data_half_court_4_zones$shot_type == 2 | data_half_court_4_zones$shot_y > 14.25 & data_half_court_4_zones$shot_type == 2, "Mid-Range", ifelse(data_half_court_4_zones$shot_distance < 4, "Restricted Area", ifelse(data_half_court_4_zones$shot_x_abs < 8 & data_half_court_4_zones$shot_y < 14.25 & data_half_court_4_zones$shot_distance > 4, "Paint (Not RA)", "Backcourt")))))

#Create dataframes based on defender distance for further analysis
`data_0-1` <- filter(data_half_court_4_zones, data_half_court_4_zones$closest_defender == "0-1")
`data_1-2` <- filter(data_half_court_4_zones, data_half_court_4_zones$closest_defender == "1-2")
`data_2-3` <- filter(data_half_court_4_zones, data_half_court_4_zones$closest_defender == "2-3")
`data_3-4` <- filter(data_half_court_4_zones, data_half_court_4_zones$closest_defender == "3-4")
`data_4-5` <- filter(data_half_court_4_zones, data_half_court_4_zones$closest_defender == "4-5")
`data_5-6` <- filter(data_half_court_4_zones, data_half_court_4_zones$closest_defender == "5-6")

#Assign names of dataframes for later parsing
Distances <- c("data_0-1", "data_1-2", "data_2-3", "data_3-4", "data_4-5", "data_5-6")

#Assign names of categories for later variable assignemnt
Zones <- c("Three", "Mid-Range", "Paint (Not RA)", "Restricted Area")

#Create new datafranes segmenting each court zone by defender range 
for (i in Distances) {
  for (j in Zones) {
  zones <- paste(i, j, sep = "_")
  d=get(i)
  assign(zones, filter(d, d$shot_zone == j))
  }
}

#Combine dataframes into one
zone_vs_distance <- ddply(data_half_court_4_zones, .(shot_zone), summarize, `0-1` = 0, `1-2` = 0, `2-3` = 0, `3-4` = 0, `4-5` = 0, `5-6` = 0)

#Assign row names
rownames(zone_vs_distance) <- c("Three", "Mid-Range", "Paint (Not RA)", "Restricted Area")

#Format
zone_vs_distance <- zone_vs_distance[,-1]

#Upload statistics from dataframes into new dataframe
zone_vs_distance$`0-1` <- c(sum(`data_0-1_Three`$made_numeric)/length(`data_0-1_Three`$made), sum(`data_0-1_Mid-Range`$made_numeric)/length(`data_0-1_Mid-Range`$made), sum(`data_0-1_Paint (Not RA)`$made_numeric)/length(`data_0-1_Paint (Not RA)`$made), sum(`data_0-1_Restricted Area`$made_numeric)/length(`data_0-1_Restricted Area`$made))

zone_vs_distance$`0-1` <- round(zone_vs_distance$`0-1`, 2)

zone_vs_distance$`1-2` <- c(sum(`data_1-2_Three`$made_numeric)/length(`data_1-2_Three`$made), sum(`data_1-2_Mid-Range`$made_numeric)/length(`data_1-2_Mid-Range`$made), sum(`data_1-2_Paint (Not RA)`$made_numeric)/length(`data_1-2_Paint (Not RA)`$made), sum(`data_1-2_Restricted Area`$made_numeric)/length(`data_1-2_Restricted Area`$made))

zone_vs_distance$`1-2` <- round(zone_vs_distance$`1-2`, 2)

zone_vs_distance$`2-3` <- c(sum(`data_2-3_Three`$made_numeric)/length(`data_2-3_Three`$made), sum(`data_2-3_Mid-Range`$made_numeric)/length(`data_2-3_Mid-Range`$made), sum(`data_2-3_Paint (Not RA)`$made_numeric)/length(`data_2-3_Paint (Not RA)`$made), sum(`data_2-3_Restricted Area`$made_numeric)/length(`data_2-3_Restricted Area`$made))

zone_vs_distance$`2-3` <- round(zone_vs_distance$`2-3`, 2)

zone_vs_distance$`3-4` <- c(sum(`data_3-4_Three`$made_numeric)/length(`data_3-4_Three`$made), sum(`data_3-4_Mid-Range`$made_numeric)/length(`data_3-4_Mid-Range`$made), sum(`data_3-4_Paint (Not RA)`$made_numeric)/length(`data_3-4_Paint (Not RA)`$made), sum(`data_3-4_Restricted Area`$made_numeric)/length(`data_3-4_Restricted Area`$made))

zone_vs_distance$`3-4` <- round(zone_vs_distance$`3-4`, 2)

zone_vs_distance$`4-5` <- c(sum(`data_4-5_Three`$made_numeric)/length(`data_4-5_Three`$made), sum(`data_4-5_Mid-Range`$made_numeric)/length(`data_4-5_Mid-Range`$made), sum(`data_4-5_Paint (Not RA)`$made_numeric)/length(`data_4-5_Paint (Not RA)`$made), sum(`data_4-5_Restricted Area`$made_numeric)/length(`data_4-5_Restricted Area`$made))

zone_vs_distance$`4-5` <- round(zone_vs_distance$`4-5`, 2)

zone_vs_distance$`5-6` <- c(sum(`data_5-6_Three`$made_numeric)/length(`data_5-6_Three`$made), sum(`data_5-6_Mid-Range`$made_numeric)/length(`data_5-6_Mid-Range`$made), sum(`data_5-6_Paint (Not RA)`$made_numeric)/length(`data_5-6_Paint (Not RA)`$made), sum(`data_5-6_Restricted Area`$made_numeric)/length(`data_5-6_Restricted Area`$made))

zone_vs_distance$`5-6` <- round(zone_vs_distance$`5-6`, 2)

#Format
zone_vs_distance <- data.matrix(zone_vs_distance)

#Create breaks for heatmap
col_breaks = c(seq(-1,0,length=100), seq(0.01,0.33,length=100), seq(0.34,0.66,length=100), seq(0.67,1,length=100)) 

#Create heatmap for field goal percentage by shot zone and defender distance
heatmap.2(zone_vs_distance, cellnote = zone_vs_distance, notecol="black", density.info="none", trace="none", margins =c(7,13.5), col="plasma", breaks=col_breaks, dendrogram="none", Colv="NA", lwid=c(.3,9), lhei=c(3,12.5), cexRow = 2, key = FALSE, main = "FG% by Shot Zone and Defender Distance", notecex = 5, xlab = "Defender Distance", Rowv=FALSE, cexCol = 2)
```

#Hypotheses and t test results
```{r}
#Effective Field Goal Percentage of Threes vs Mid_Range Shots
t.test(filter(data_half_court, data_half_court$shot_type == 3)$made_numeric*1.5, filter(data_half_court, data_half_court$shot_zone == "Mid-Range")$made_numeric, conf.level = .99, alternative = 'greater')

#Field Goal Percentage of Contested Layups vs Open Mid-Range Shots
t.test(filter(data_half_court, data_half_court$defender_distance < 2 & data_half_court$shot_zone == "Restricted Area")$made_numeric, filter(data_half_court, data_half_court$defender_distance > 6 & data_half_court$shot_zone == "Mid-Range")$made_numeric, conf.level = .99, alternative = 'greater')

#Field Goal Percentage of threes contested within .5-2.5 feet vs 2.5-4.5 feet
t.test(filter(data_half_court, data_half_court$defender_distance > 2.5 & data_half_court$defender_distance < 4.5 & data_half_court$shot_type == 3)$made_numeric, filter(data_half_court, data_half_court$defender_distance > .5 & data_half_court$defender_distance < 2.5 & data_half_court$shot_type == 3)$made_numeric, conf.level = .99, alternative = 'greater')

#Field Goal Percentage of Pull Up Threes vs Catch and Shoot Threes
t.test(filter(data_half_court, data_half_court$dribbles_before == 0 & data_half_court$shot_type == 3)$made_numeric, filter(data_half_court, data_half_court$dribbles_before > 0 & data_half_court$shot_type == 3)$made_numeric, conf.level = .99, alternative = 'greater')

#Field Goal Percentage of threes while moving fast vs more stationary threes
t.test(filter(data_half_court, data_half_court$shooter_velocity_ft_sec < 9 & data_half_court$shot_type == 3 & data_half_court$shot_distance < 27)$made_numeric, filter(data_half_court, data_half_court$shooter_velocity_ft_sec > 9 & data_half_court$shot_type == 3 & data_half_court$shot_distance < 27)$made_numeric, conf.level = .99, alternative = 'greater')

#Field Goal Percentage of layups contested within 2-4 feet vs 4-6 feet
t.test(filter(data_half_court, data_half_court$defender_distance > 4 & data_half_court$defender_distance < 6, data_half_court$shot_zone == "Restricted Area")$made_numeric, filter(data_half_court, data_half_court$defender_distance > 2 & data_half_court$defender_distance < 4 & data_half_court$shot_zone == "Restricted Area")$made_numeric, conf.level = .99, alternative = 'greater')

#Effective Field Goal Percentage of open catch and shoot threes vs contested drives to the paint
t.test(filter(data_half_court, data_half_court$dribbles_before == 0 & data_half_court$defender_distance > 5 & data_half_court$shot_type == 3)$made_numeric*1.5, filter(data_half_court, data_half_court$dribbles_before > 0 & data_half_court$defender_distance < 2 & data_half_court$shot_zone == "Paint (Not RA)")$made_numeric, conf.level = .99, alternative = 'greater')
```

#Naive Bayes model to predict probability of a made shot
```{r}
#New dataframe from previous data
data_10_zones <- data

#Create ten different shot zones
data_10_zones$shot_zone <- as.factor(ifelse(data_10_zones$shot_type == 3 & data_10_zones$shot_y < 9.25 & data_10_zones$shot_x > 0, "Right Corner Three", ifelse(data_10_zones$shot_type == 3 & data_10_zones$shot_y < 9.25 & data_10_zones$shot_x < 0, "Left Corner Three", ifelse(data_10_zones$shot_type == 3 & data_10_zones$shot_y > 9.25 & data_10_zones$shot_y < 42.25 & data_10_zones$shot_distance < 27 & data_10_zones$shot_distance > 6, "Above the Break Three", ifelse(data_10_zones$shot_x_abs > 8 & data_10_zones$shot_type == 2 & data_10_zones$shot_distance < 20 | data_10_zones$shot_y > 14.25 & data_10_zones$shot_type == 2 & data_10_zones$shot_distance < 20 , "Mid-Range", ifelse(data_10_zones$shot_x_abs > 8 & data_10_zones$shot_type == 2 | data_10_zones$shot_y > 14.25 & data_10_zones$shot_type == 2, "Deep Two", ifelse(data_10_zones$shot_distance < 4, "Restricted Area", ifelse(data_10_zones$shot_x_abs < 8 & data_10_zones$shot_y < 14.25 & data_10_zones$shot_distance > 4 & data_10_zones$shot_distance < 6, "Close Paint (Not RA)", ifelse(data_10_zones$shot_x_abs < 8 & data_10_zones$shot_y < 14.25 & data_10_zones$shot_distance > 4 & data_10_zones$shot_distance > 6, " Deep Paint (Not RA)", ifelse(data_10_zones$shot_type == 3 & data_10_zones$shot_y > 9.25 & data_10_zones$shot_y < 42.25 & data_10_zones$shot_distance > 27, "Deep Three", "Backcourt"))))))))))

#5 fold cross validation with 5 repeats
for (i in 1:5) {
folds <- createFolds(data_10_zones$made, k = 5)

train1 <- data_10_zones[c(folds[[2]], folds[[3]], folds[[4]], folds[[5]]), ]

train2 <- data_10_zones[c(folds[[1]], folds[[3]], folds[[4]], folds[[5]]), ]

train3 <- data_10_zones[c(folds[[2]], folds[[1]], folds[[4]], folds[[5]]), ]

train4 <- data_10_zones[c(folds[[1]], folds[[3]], folds[[2]], folds[[5]]), ]

train5 <- data_10_zones[c(folds[[2]], folds[[1]], folds[[3]], folds[[4]]), ]

test1 <- data_10_zones[folds[[1]], ]

test2 <- data_10_zones[folds[[2]], ]

test3 <- data_10_zones[folds[[3]], ]

test4 <- data_10_zones[folds[[4]], ]

test5 <- data_10_zones[folds[[5]], ]

nb_class_1 <- naiveBayes(made ~ defender_distance + shot_zone, data = train1)

nb_predictions_1 <- predict(nb_class_1, test1, type = 'class')

accuracy_1 <- mean(nb_predictions_1 == test1$made)

nb_class_2 <- naiveBayes(made ~ defender_distance + shot_zone, data = train2)

nb_predictions_2 <- predict(nb_class_2, test2, type = 'class')

accuracy_2 <- mean(nb_predictions_2 == test2$made)

nb_class_3 <- naiveBayes(made ~ defender_distance + shot_zone, data = train3)

nb_predictions_3 <- predict(nb_class_3, test3, type = 'class')

accuracy_3 <- mean(nb_predictions_3 == test3$made)

nb_class_4 <- naiveBayes(made ~ defender_distance + shot_zone, data = train4)

nb_predictions_4 <- predict(nb_class_4, test4, type = 'class')

accuracy_4 <- mean(nb_predictions_4 == test4$made)

nb_class_5 <- naiveBayes(made ~ defender_distance + shot_zone, data = train5)

nb_predictions_5 <- predict(nb_class_5, test5, type = 'class')

accuracy_5 <- mean(nb_predictions_5 == test5$made)

accuracy.rating <- paste("accuracy", i, sep = ".")
assign(accuracy.rating, (accuracy_1 + accuracy_2 + accuracy_3 + accuracy_4 + accuracy_5)/5)
}
#Accuracy of model
(accuracy.1 + accuracy.2 + accuracy.3 + accuracy.4 + accuracy.5)/5